The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-01 manuscript/manuscript.Rnw
add changes that amy proposed

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
spanning the phylogenetic tree of mammals. This revealed a dynamic
expression pattern of expression of lineage-specific expansions of
gene families and differential gene expression that agreed with the
scenario of an “arms race” between maternal and fetal tissues that
have shaped such disparate gene expression profiles. In addition, we
highlighted the broad spectrum of hormone gene expression in multiple
species that suggested that there is a lack of conservation in hormone
signaling during mammalian pregnancy. Major differences between
expression in marsupial and eutherian mammals at term seemed to be
correlated with the differences between choriovitelline and
chorioamniotic placenta, with the marsupial placenta more enriched in
genes the contributed to the active generation of energy. We also
identified that galectin expression, especially that of
\textit{LGALS1}, is a common feature of all mammalian placentas and
was most likely present in the ancestral therian. Also, we identified
two new expansions of genes that are highly expressed in the \elephant
placenta hat may be involved in trophoplast differentiation and immune
system function, mirroring expansions in other groups such as
primates, rodents, and ruminants. We also quantified significant
lineage-specific shifts in gene expression, concentrating on the
identification of genes that showed gene expression changes on the
human lineage. We cautiously propose that many of these genes may be
related to the invasive nature of the trophoblast in \human, although
we note that in many of these genes, there is in some cases
substantial expression in the \bonobo as well. These genes have the
potential to spur further research and may be at the root of
obstetrical disorders such as pre-eclampsia. We propose that more
placentas from diverse lineages, especially those of marsupials and
primates, to further identify genes and expression levels that are
specific to marsupials and eutherians, as well as confirm the identity
of human-specific expression shifts.

@@

  Jason Caravas provided samples.

>>>>

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a cesarean section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

\DLA{Totally rewrite the conclusion; it's terrible}

We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
spanning the phylogenetic tree of mammals. This revealed a dynamic
expression pattern of expression of lineage-specific expansions of
gene families and differential gene expression that agreed with the
scenario of an “arms race” between maternal and fetal tissues that
have shaped such disparate gene expression profiles. In addition, we
highlighted the broad spectrum of hormone gene expression in multiple
species that suggested that there is a lack of conservation in hormone
signaling during mammalian pregnancy. Major differences between
expression in marsupial and eutherian mammals at term seemed to be
correlated with the differences between choriovitelline and
chorioamniotic placenta, with the marsupial placenta more enriched in
genes which contribute to the active generation of energy. We also
identified that galectin expression, especially that of
\textit{LGALS1}, is a common feature of all mammalian placentas and
was most likely present in the ancestral therian. Also, we identified
two new expansions of genes that are highly expressed in the \elephant
placenta that may be involved in trophoplast differentiation and immune
system function, mirroring expansions in other groups such as
primates, rodents, and ruminants. We also quantified significant
lineage-specific shifts in gene expression, concentrating on the
identification of genes that showed gene expression changes on the
human lineage. We cautiously propose that many of these genes may be
related to the invasive nature of the trophoblast in \human, although
we note that in many of these genes, there is in some cases
substantial expression in the \bonobo as well. These genes have the
potential to spur further research and may be at the root of
obstetrical disorders such as pre-eclampsia. We propose that more
placentas from diverse lineages, especially those of marsupials and
primates, to further identify genes and expression levels that are
specific to marsupials and eutherians, as well as confirm the identity
of human-specific expression shifts.

@@

  Jason Caravas performed initial analyses.
